---
title: "Cluster Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

Reading in the data and doing an example of Cluster Analysis

```{r}
drugs <- read.csv(file.choose(),header = TRUE)
#Deleting a column Brant added by accident
drugs <- drugs[,-1]
#Making state names the row names
row.names(drugs) <- drugs$State
#making a matrix
drugsMatrix <- as.matrix(drugs[,2:6])

#Scaling the matrix for the distance function
drugsMatrix <- scale(drugsMatrix)

#calculating the initial distance matrix
drugsMatrix.dist <- dist(drugsMatrix, method="euclidean")

#Performing the analysis with the single linkage method
drugs.clust <- hclust(drugsMatrix.dist, method = "single")

#Plotting the result 
#Interpretation note: height is the distance that the clusters fused at, given the method we chose
plot(drugs.clust)

#Performing the analysis with the complete method
drugs.clust2 <- hclust(drugsMatrix.dist, method = "comp")

#Plotting the result
plot(drugs.clust2)

#Performing the analysis with the ward method
drugs.clust3 <- hclust(drugsMatrix.dist, method = "ward.D")

#Plotting the result
plot(drugs.clust3)

#The function to extract clusters is called cutree and will take a height or a number of groups
clusters.ward <- cutree(drugs.clust3, k = 6)
head(clusters.ward)

#Brant's function to create parallel plot, clust.grps (need to load MASS package for this to work)
clust.grps = function(X,grps,parcoord=F,suppress=F) {
    k = length(unique(grps))
    p = dim(X)[[2]]
    Xmeans = matrix(0,nrow=length(unique(grps)),ncol=p+1)
    X = as.data.frame(X)
    for (i in 1:k){
        cat("\n")
        cat(paste("Cluster",i,"\n"))
        cat("=======================================================================\n")
        if (suppress==F){
          cat(row.names(X)[grps==i])
          cat("\n\n")}
        cat("Variable means in this cluster are:\n")
        cat("----------------------------------------------------------------------\n")
        print(apply(X[grps==i,],2,mean))
        Xmeans[i,]=c(apply(X[grps==i,],2,mean),as.numeric(i))
        cat("\n\n")
        
    } 
    if (parcoord) {parcoord(Xmeans[,-(p+1)],col=as.numeric(Xmeans[,(p+1)])+3,
                            lwd=2,var.label=T)} 
}

#using this function
#input: data.table of data that went into cluster analysis and the output of the cutree function
clust.grps(drugs[,2:6],clusters.ward, parcoord = TRUE)

```
Some notes on outliers:
Outliers really gum up the works here because extreme cases may end up in clusters on their own when running the analysis.To dampen the effect of an outlier, you can use the "taxi cab" distance instead of the scaled euclidean. Also, sometimes okay to remove the outliers, depending on the analysis you're performing. Also, this is one of many methods that can be used to discover outliers.

###MONDAY AFTERNOON WORK




